GPU-friendly Stroke Expansion

### 2024-04-12 paper.tex
First half of the GPU implementation section
- Started Section 8. Please look at sections 8 and 8.1.
- I moved the existing input and path encoding drafts into section 8.
  Some changes to this will follow.
- I put down a lot of TODOs in the text that can use your feedback. Some
  of them are for tasks for next week, others are genuine questions.
- Stubs for incomplete Section 8 subsections.

\documentclass[format=acmsmall]{acmart}
\begin{document}
\title{GPU-friendly Stroke Expansion}
\author{Raph Levien}
\author{Arman Uguray}
\affiliation{
    \institution{Google}
    \city{San Francisco}
    \state{CA}
    \country{USA}
}

@@

TODO: discussion of prior art.

@@

\section{Input encoding}

@@

@@

>>>>

\documentclass[format=acmsmall]{acmart}
\usepackage[linesnumbered]{algorithm2e}

\begin{document}
\title{GPU-friendly Stroke Expansion}
\author{Raph Levien}
\author{Arman Uguray}
\affiliation{
    \institution{Google}
    \city{San Francisco}
    \state{CA}
    \country{USA}
}

@@

TODO: the introduction should summarize the paper's contributions and the outline of the rest of the paper.
TODO: discussion of prior art.

@@

@@

\section{GPU Implementation}

We applied the techniques outlined in Section 7 to implement a data-parallel algorithm that can convert stroked 2D Bézier paths into flat geometry suitable for GPU rendering. We focused primarily on the global stroke-to-fill conversion problem in order to generate polygonal outlines for rasterization. Since the Euler spiral approximation can fit any source cubic Bézier and its parallel offset curves, our implementation can be used to render both filled and stroked primitives within a satisfactory error tolerance.

We aimed to satisfy the following the criteria:
\begin{enumerate}
  \item The implementation must be able to handle a large number of inputs at real-time rates. 
  \item The stroke outlines must meet the weak correctness definition.
  \item The implementation must support the standard SVG stroke end cap (\emph{butt}, \emph{square}, \emph{round}) and join styles (\emph{bevel}, \emph{miter}, \emph{round}).
  \item The CPU must do no work beyond the basic preperation of the input path data for GPU consumption.
\end{enumerate}

Our implementation consists of a single GPU compute shader that satisfies these criteria. We coupled this with a simple and efficient data layout for parallel processing. Notably, our implementation is fully data-parallel on input path segments and does not require any computationally expensive curve evaluation on the CPU. We implemented our shader on general-purpose GPU compute primitives supported by all modern graphics APIs, particularly Metal, Vulkan, D3D12, and WebGPU[? citation?]. As such, the ideas presented in this section are easily portable.

The rest of this section describes the design of our GPU pipeline and input encoding scheme.

% TODO: WGSL, minimal compute requirements
% TODO: workgroup layout/dimensions, parallelism
% TODO: - impl details unordered across shapes. This must also be highlighted.
%       - Watertightness
%       - error tolerance used.
% TODO: copy other TODOs
% TODO: dashing as future work

\subsection{Pipeline Design}

An SVG path consists of a sequence of instructions (or "verbs"). The \emph{lineto} and \emph{curveto} instructions denote an individual \emph{path segment} consisting of either a straight line or a cubic Bézier. A given subsequence of path segments can be bounded by a \emph{moveto} and/or \emph{closepath} instruction which mark the beginning and end of a \emph{subpath}. A subpath must form a closed contour if painted as a fill. When painted as a stroke, each subpath must begin and end with a cap according to the desired cap style. In addition, each path segment in a stroked subpath must be connected to adjacent path segments of the same subpath with a join. For any given path, the cap and join styles do not vary across subpaths or individual path segments.

The standard organizational primitive for a compute shader is the \emph{workgroup}. A workgroup can be organized as a 1, 2, or 3-dimensional grid of individual threads that can cooperate using shared memory. Multiple workgroups can be \emph{dispatched} as part of a larger global grid, also of 1, 2, or 3 dimensions.

Our compute shader is arranged as a 1-dimensional grid, parallelized over input path segments. We chose a workgroup size of 256 as it works well over a wide range of GPUs. Each thread in the global grid is assigned to process a single path segment. We encode the individual path segments contiguously in a GPU storage buffer such that each element has 1:1 correspondance to a global thread ID in the dispatch. This layout easily scales to tens of thousands of input path segments, which can be distributed across any number of paths and subpaths.

% TODO: caps/joins figure highlighting mapping of segments, joins, and caps to threads, encoding.
% TODO: visualize per-segment flattened output for a fill and for a stroke from the perspective of one thread.
% TODO: finish this \begin{algorithm} snippet
\begin{algorithm}
  $tid \gets$ global invocation ID\;
  $(style, segment) \gets readScene(tid)$\;
  \eIf{$style$ is $fill$}{
      flatten source curve
  }{\If{you're sad}{
      just keep going
  }}
\caption{The control flow of our compute shader}
\end{algorithm}

Each thread outputs a polyline that fits a subset of the rendered subpath's outer contour, represented by the path segment. For a filled primitive, a thread only outputs the flattened approximation of the path segment. For a stroke, a thread outputs the polylines approximating multiple curves: a) the two parallel curves on both sides of the source segment, offset by the desired stroke \emph{halfwidth}, and b) the inner and outer joins that connect the path segment to the next adjacent segment. We also store metadata in our input encoding that marks the position of a path segment within the containing subpath. We use this information to determine whether to output an end cap instead of a join. We encode one additional segment designated as the \emph{stroke cap marker} for every subpath. The thread assigned to the marker segment only outputs a start cap, which completes the stroke outline.

The joins between adjacent segments need to form a continuous outline when combined with the segments. This requires that a thread have access to the tangent vector at the end of its assigned segment \emph{and} the start tangent of the next adjacent segment. We require that the input segments within a subpath are stored in order which simplifies the access to the adjacent segment down to a simple read at the next array offset. Co-operation across threads is not necessary.

A property of this data-parallel structure is that the output is unordered. The cubic-Bézier-to-Euler-spiral conversion is sequential within a single thread, so it is possible to preserve ordering at the scope of an individual path segment, if desired. However, ordering cannot be guaranteed across path segments. We refer to this output data structure as \emph{line soup}. We believe this structure can be adapted to other primitives, such as circular arc segments or triangles. We incorporated our compute shader into our own massively-parallel GPU rasterizer that spatially sorts the line soup in subsequent binning and tiling stages. As such, maintaining any sort of ordering with respect to the input path data structure was not necessary.

% TODO: Talk about the inner join here. We detect inner vs outer outline based on join tangent cross product. We just draw a straight line for the inner join.
% TODO: line soup is bump allocated. Mention this here? I'm leaning towards discussing that in the Limitations section below, which will focus on memory allocation and estimation.
% TODO: while parallel on input, sequential per curve. However iterations are bounded (? needs measurement). Should this discussion happen here or down in the Performance section?

\subsection{Input encoding}

% TODO: not sure how much to emphasize the path tag encoding. The input processing monoids benefit the vello scene API (fragments) the TagMonoid could generally be computed on the CPU (with minimal offset tracking) It's not that relevant to the stroke processing. The main bit to mention is that it's a struct-of-arrays.
% TODO: talk about sanitization of input: closepaths are handled ahead of time, stroke cap marker.
% TODO: discuss style encoding in greater detail

\subsection{Algorithm}

% TODO: explain flatten_euler

\subsection{Limitations}

% TODO: memory usage discussion: the impact of lines and how arcs might help. or work graphs, or maybe mesh shaders. Limitation: memory must be allocated up-front. bump estimation, violates the no pre-curve-evaluation criterium but simpler techniques (like Wang's) exist and we use it but way over-estimates. This is a limitation of the GPU execution model. Explain this trade-off.
  % -> TODO: Memory usage graphs for input scene buffers and LineSoup.

\section{Performance}

% TODO: We integrated our algorithm into a fully GPU-driven rendering pipeline, Vello WGSL
% TODO: 3 alternative implementations: cubic->ES (for load balancing eval), cubic->ES->LineSoup (current), cubic->ES->ArcSoup
% TODO: GPUs tested, data sets. Test strategy
% TODO: divergence (reference prim count)

TODO

@@

% TODO: mention dashing as future work?
